---
title: "Enterprise Mirror"
vignette: >
  %\VignetteIndexEntry{Enterprise Mirror}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Some networks never reach the internet, and some teams approve packages before
anyone installs them. `collection_mirror()` serves both. It builds an R console
backed by its own package repository. Visitors can `install.packages()`
anything in that repository and nothing from outside it, because a mirror is
always offline.

![A mirror answers `install.packages()` from its own folder. The call reads the mirror's index (1), then installs the package beside it (2).](images/mirror.svg){#fig-mirror fig-alt="A visitor's console runs install.packages for dplyr. In the mirror folder's file tree, the request goes first to the PACKAGES index under repo/bin/emscripten/contrib/4.6/ and then to the dplyr file beside it. The folder also holds the console page, the webR engine, LICENSES/, mirror-config.json and _headers, and nothing is fetched from outside it"}

A mirror differs from a `bind()` site in what goes in and in who installs the
packages.

| | `bind()` | `collection_mirror()` |
|---|---|---|
| Input | a collection with `_webrarian.yml` | a list of packages, or whole repositories |
| Your files and scripts | bundled | none |
| Packages | installed when the page opens | installed by visitors with `install.packages()` |
| Requests to other servers | none with `build.offline: true` | none, ever |
| Share links | `"open"` by default | `"off"` by default |

## Building a mirror

Start with an empty console, name the packages, or take everything.

```{r}
library(webrarian)

# An empty R console
collection_mirror("webr-mirror")

# Packages visitors may install, with their dependencies
collection_mirror(
  "webr-mirror",
  packages = c("dplyr", "ggplot2", "specialpackage"),
  repos = "https://myorg.r-universe.dev"
)

# Every package of repo.r-wasm.org and of `repos`: tens of gigabytes
collection_mirror("/srv/webr-mirror", mode = "full")
```

Packages are looked up in repo.r-wasm.org, then each of `repos`. Downloads are
checked against the repository's checksums, and a package that cannot be
obtained stops the build. `mode = "full"` answers the air gap and rather
defeats the approval list.

Three other arguments are worth knowing. `webr_version` pins the engine (a
tested version, or a newer patch release with a warning), `favicon` sets the
icon, and `share_links` decides what share links may do
([Customization](customization.html#share-links)).

The mirror is one directory.

```text
webr-mirror/
├── index.html              the console page
├── exlibris-r.<hash>.js    the viewer
├── exlibris-r.<hash>.css
├── webr/v0.6.0/            the webR engine
├── repo/bin/emscripten/contrib/4.6/
│   ├── dplyr_<version>.tgz
│   └── PACKAGES, PACKAGES.gz, PACKAGES.rds
├── LICENSES/               licenses of the engine, the viewer and every package
├── sw.js                   removes any earlier service worker at this address
├── mirror-config.json      what was mirrored, and when
└── _headers                response headers for hosts that read them
```

To update, run the same call again. Files whose checksums match are kept and
the index is rewritten. An interrupted build resumes the same way.

## Serving the mirror

Serve the directory with the two cross-origin isolation headers
([Deployment](deployment.html#cross-origin-isolation)). Without them Ctrl+C
cannot interrupt R. `reading_room("webr-mirror")` serves it locally with the
right headers. Here is the configuration for nginx.

```nginx
server {
    listen 443 ssl;
    server_name webr.internal.example.com;
    ssl_certificate     /etc/ssl/certs/internal.crt;
    ssl_certificate_key /etc/ssl/private/internal.key;
    root /srv/webr-mirror;

    add_header Cross-Origin-Opener-Policy "same-origin" always;
    add_header Cross-Origin-Embedder-Policy "require-corp" always;
    add_header Cache-Control "no-cache" always;

    location ~ ^/(webr/|exlibris-r\.) {
        add_header Cross-Origin-Opener-Policy "same-origin" always;
        add_header Cross-Origin-Embedder-Policy "require-corp" always;
        add_header Cache-Control "public, max-age=31536000, immutable" always;
    }
}
```

For Apache, the two isolation headers are set as follows.

```apache
<Directory /srv/webr-mirror>
    Header always set Cross-Origin-Opener-Policy "same-origin"
    Header always set Cross-Origin-Embedder-Policy "require-corp"
</Directory>
```

## Air-gapped networks

Four steps bring a mirror into a closed network.

1. Build the mirror on a machine with internet access.
2. Move the directory into the closed network by your approved method.
3. Serve it from an internal web server.
4. Open the page and run `install.packages("dplyr")` to confirm it installs
   from the mirror.

The page sets webR's `webr_pkg_repos` option to the mirror's `repo/` only, so
the console, the Packages tab and share links install from there.
`getOption("repos")` is unchanged, so `available.packages()` does not see the
mirror.

If an install fails, check that the package is in
`repo/bin/emscripten/contrib/4.6/PACKAGES`, look for 404s in the browser's
developer tools, and make sure the server does not rewrite files under
`repo/`. With a self-signed certificate, add your certificate authority to the
each browser's trust store.
