---
title: "Bundling Files"
vignette: >
  %\VignetteIndexEntry{Bundling Files}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

R in the browser cannot read the visitor's disk, which is the browser working
as intended. It sees a virtual file system instead, and that file system
starts empty. `bind()` copies each file you include into the site, and the
page copies each one into the file system when it opens.

![`bind()` copies each bundled file into `vfs-files/`, and the page copies it into the browser's file system at startup.](images/vfs-files.svg){#fig-vfs fig-alt="Three places hold the same two files. analysis.R and data/survey.csv in the collection, drawn as a project folder, become vfs-files/analysis.R and vfs-files/data/survey.csv in the site, drawn as a server, and /home/web_user/analysis.R and /home/web_user/data/survey.csv in the browser, drawn as a window, where R starts"}

Each file keeps its path, so `data/survey.csv` becomes
`_site/vfs-files/data/survey.csv` in the site and
`/home/web_user/data/survey.csv` in the browser. The page requests each file
separately. There is no archive to unpack.

## Adding files

`acquire_file()` adds paths, directories and glob patterns to
`files.include`.

```{r}
library(webrarian)

acquire_file("data/")
acquire_file("scripts/*.R")
acquire_file(c("config/*.yml", "assets/"))
```

Each pattern must match at least one existing file inside the collection.
Otherwise `acquire_file()` refuses it and leaves `_webrarian.yml` unchanged.

### Pattern rules

The table shows what each pattern selects.

| Pattern | Selects |
|---------|---------|
| `data/` | every file under `data/`, at any depth, except dot files |
| `data/*` | the files directly inside `data/` |
| `*.R` | the `.R` files at the top of the collection |
| `**/*.csv` | the `.csv` files at any depth, the top level included |
| `*.{csv,rds}` | the `.csv` and `.rds` files at the top |
| `.` | every file in the collection, except dot files and dot directories |
| `.Rprofile` | that dot file |
| `.github/` | every file under that dot directory, except dot files inside it |

Patterns are relative to the collection and select files, never bare
directories.

A file or directory whose name starts with a dot is selected only by a
pattern that spells the dot. The patterns `.`, `**`, `data/` and `*.RData` all
skip
`.Renviron`, `.env`, `.RData` and anything inside a dot directory. A broad
include therefore cannot pick up `.Renviron` or `.env` by accident. A pattern
has to spell the dot to select them.

## Leaving files out

`files.exclude` takes `.gitignore`-style patterns and removes files the
include patterns selected.

```yaml
# _webrarian.yml
files:
  include:
    - "data/"
  exclude:
    - "data/raw/"
    - "*.log"
```

A pattern without a slash (`*.log`) matches at any depth, a trailing slash
(`data/raw/`) leaves out a directory, and a slash inside (`data/secret.csv`)
anchors it at the collection root. The defaults are `**/*.Rhistory` and
`**/.DS_Store`, and `exclude: []` removes them.

The output directory, `.webrarian/`, `.git/`, `.Rproj.user/`, `renv/library/`,
`renv/staging/`, `node_modules/` and `_webrarian.yml` are never bundled,
whatever the patterns. A symbolic link leading outside the collection is skipped
with a warning.

## Checking and removing

See what `bind()` will copy, or take a pattern back out.

```{r}
collection_files()$files   # what bind() will copy
withdraw_file("data/temp/")
```

`bind()` also warns about any include pattern that matches nothing.

## File locations

Files appear under the mount point, `/home/web_user` by default, and R starts
there, so paths work as they do on your machine.

```r
survey <- read.csv("data/survey.csv")
source("scripts/helpers.R")
```

After a `setwd()`, use the full path (`/home/web_user/data/survey.csv`). The
mount point is itself a setting.

```{r}
settings_set("files.mount-point" = "/project")
```

## File size

Every visitor downloads every bundled file when the page opens. Prefer RDS to
large CSV files, a sample to the full dataset, and the files the analysis
reads to the files you happen to have.

## Running scripts when the page opens

Three `repl` settings name bundled files by their path in the collection.

- `startup-script` runs once, after the files are in place and before anything
  opens. It also runs for visitors arriving through a share link, and a link
  cannot replace it. A shared file at its path is saved beside it as
  `<name>-shared.<ext>`.
- `auto-open` lists files to open in the editor. Unset, the first bundled R
  file other than the startup script opens, and `auto-open: []` opens none.
- `auto-run` lists files to run, in order, once the console is ready.

![The page works through its startup in a fixed order. It starts webR (1), installs packages (2) and places files (3), then runs `startup-script` (4), opens the `auto-open` files (5) and runs the `auto-run` files (6).](images/startup.svg){#fig-order fig-alt="Six close-ups of the workspace panes, in order, each with a small map of the pane it shows. Loading R, the console installing packages, the Files tab with the bundled files, the startup script's objects in the Environment tab, a file open in the editor for repl.auto-open, and the plot from the repl.auto-run code"}

A collection might set them as follows.

```yaml
# _webrarian.yml
files:
  include:
    - "data/"
    - "scripts/"

repl:
  startup-script: "scripts/setup.R"
  auto-open:
    - "scripts/analysis.R"
  auto-run: []
```

`bind()` stops with an error when an entry is not a bundled file. A bare file
name works when exactly one bundled file has that name. A startup script can
be as short as this one.

```r
# scripts/setup.R
library(dplyr)
survey <- readRDS("data/survey.rds")
message("Ready: `survey` is loaded.")
```

## Missing files

Work through these checks when R cannot find a file.

1. Run `getwd()` in the browser, because code that called `setwd()` may have
   moved R away from the mount point.
2. Check `collection_files()$files`. A dot file needs a pattern that names
   the dot.
3. Look in `_site/vfs-files/`, which holds exactly what the page copies.
4. Run `bind()` again after adding files.
